Refactoring the `nls()` function in R

### 2022-06-15 RefactoringNLS/RefactoringNLS.Rmd
Minor edits
Continued work on Refactoring

This article considers the features and limitations of the R function `nls()`. It
arose from a Google Summer of Code project "Improvements to `nls()`", a tool 
intended for the estimation of models written as a formula that has at least
one parameter that is "nonlinear". That is, the model is not estimable via solving a 
finite set of 
linear equations. A companion document **Variety in Nonlinear Least Squares Codes** 
presents an overview of methods for the problem which takes a much wider view of the
problem of minimizing a function that can be written as a sum of squared
terms, and of the even more general case of nonlinear regression.

An important overall consideration in our work has been the maintainability of
the code base that supports the `nls()` functionality, as we believe that the
existing code makes maintenance and improvement very difficult. Moreover, legacy
applications and examples mean some improvements are blocked. Thus we have not
addressed all the issues raised, but we believe that the discussion and trials
so far are relevant to improving both `nls()` and R generally.

@@

`nls()` is the tool in base R (the distributed software package from [CRAN](https://cran.r-project.org)
for estimating nonlinear statistical models. The function was developed mainly in the 1980s and
1990s by Doug Bates et al., initially for S (see https://en.wikipedia.org/wiki/S_%28programming_language%29) . 
The ideas spring primarily from the book by @bateswatts.

The `nls()` function has a remarkable and quite comprehensive set of capabilities for estimating
nonlinear models that are expressed as formulas. In particular, we note that it

- handles formulas that include R functions
- allows data to be subset
- permits parameters to be indexed over a set of related data
- produces measures of variability (i.e., standard error estimates) for the estimated parameters
- has related profiling capabilities for exploring the likelihood surface as parameters are changed

With such a range of features and a long history, it is not surprising that the code has become untidy 
and overly patched. It is, to
our mind, essentially unmaintainable. Moreover, its underlying methods can and should be improved. Let
us review some of the issues. We will then propose corrective actions, some of which we have carried out
either in demonstrations or in the distributed code.

@@

This amounts to admitting R cannot solve perfectly well-posed problems. The 
suggestion that one needs to pollute data with errors is one that should be offensive
to good science.

It turns out that this issue can be easily resolved. The key "convergence test" -- more properly
a "termination test" for the **program** rather than for convergence of the underlying **algorithm** --
is the Relative Offset Convergence Criterion (see @BatesWatts81). This works by projecting the proposed
step in the parameter vector on the gradient and estimating how much the sum of squares loss function 
will decrease. This is divide by the current size of the loss function to avoid scale issues.
When we have "converged", the estimated
decrease is very small, as usually is its ratio to the sum of squares. However, with small residuals,
the sum of squares loss function is (almost) zero and we get the possibility of a 
zero-divide failure.

@@

### Example of a small-residual problem

```
t <- -10:10
y <- 100/(1 + .1 * exp(-0.51 * t))
lform <- y ~ a/(1 + b * exp(-c * t))
ldata <- data.frame(t = t, y = y)
plot(t, y)
lstartbad <- c(a = 1, b = 1, c = 1)
lstart2 <- c(a = 100, b = 10, c = 1)
nlsr::nlxb(lform, data = ldata, start = lstart2)
nls(lform, data = ldata, start = lstart2, trace = TRUE)
# Fix with scaleOffset
nls(lform, data = ldata, start = lstart2, trace = TRUE, control = list(scaleOffset = 1.0))
sessionInfo()
```

Edited output of running this function follows:

```
> t <- -10:10
> y <- 100/(1 + .1 * exp(-0.51 * t))
> lform <- y ~ a/(1 + b * exp(-c * t))
> ldata <- data.frame(t=t, y=y)
> plot(t,y)
> lstartbad <- c(a = 1, b = 1, c = 1)
> lstart2 <- c(a = 100, b = 10, c = 1)
> nlsr::nlxb(lform, data = ldata, start = lstart2)
nlsr object: x 
residual sumsquares =  1.007e-19  on  21 observations
    after  13    Jacobian and  19 function evaluations
  name            coeff          SE       tstat      pval      gradient    JSingval   
a                    100     2.679e-11  3.732e+12  1.863e-216  -6.425e-11       626.6  
b                    0.1      3.78e-13  2.646e+11  9.125e-196  -3.393e-08       112.3  
c                   0.51       6.9e-13  7.391e+11  8.494e-204   1.503e-08       2.791  
> nls(lform, data = ldata, start = lstart2, trace = TRUE)
40346.    (1.08e+00): par = (100 10 1)
11622.    (2.93e+00): par = (101.47 0.49449 0.71685)
5638.0    (1.08e+01): par = (102.23 0.38062 0.52792)
642.08    (1.04e+01): par = (102.16 0.22422 0.41935)
97.712    (1.79e+01): par = (100.7 0.14774 0.45239)
22.250    (1.78e+02): par = (99.803 0.093868 0.50492)
0.025789  (1.33e+03): par = (100.01 0.10017 0.50916)
6.0571e-08 (7.96e+05): par = (100 0.1 0.51)
4.7017e-19 (1.86e+04): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
... (approx 40 lines omitted)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
Error in nls(lform, data = ldata, start = lstart2, trace = TRUE) : 
  number of iterations exceeded maximum of 50
> # Fix with scaleOffset
> nls(lform, data = ldata, start = lstart2, trace = TRUE, control = list(scaleOffset = 1.0))
40346.      (1.08e+00): par = (100 10 1)
11622.      (2.91e+00): par = (101.47 0.49449 0.71685)
5638.0      (9.23e+00): par = (102.23 0.38062 0.52792)
642.08      (5.17e+00): par = (102.16 0.22422 0.41935)
97.712      (2.31e+00): par = (100.7 0.14774 0.45239)
22.250      (1.11e+00): par = (99.803 0.093868 0.50492)
0.025789    (3.79e-02): par = (100.01 0.10017 0.50916)
6.0571e-08  (5.80e-05): par = (100 0.1 0.51)
4.7017e-19  (1.62e-10): par = (100 0.1 0.51)
Nonlinear regression model
  model: y ~ a/(1 + b * exp(-c * t))
   data: ldata
     a      b      c 
100.00   0.10   0.51 
 residual sum-of-squares: 4.7e-19
Number of iterations to convergence: 8 
Achieved convergence tolerance: 1.62e-10

Running under: Linux Mint 20.3
```

@@

The single convergence criterion of `nls()` leaves out some possibilities that could
be useful for some problems. The package `nlsr` (@nlsr-manual) already offers both 
the safeguarded relative offset test (**roffset**) as well as a **small sum of 
squares** test (**smallsstest**) that compares the latest evaluated sum of squared
(weighted) residuals to `e4` times the initial sum of squares. The value of 
`e4` is computed as

```{r meps4, echo=TRUE}
epstol<-100*.Machine$double.eps
e4 <- epstol^4
e4
```

We note that `nls()` uses a termination test to stop after `maxiter` "iterations". 
Unfortunately, for almost all
iterative algorithms, the meaning of "iteration" requires careful examination of the
code. we prefer to express such tests using the number of times the residuals or the jacobian have
been computed and put upper limits on these. Our codes exit (terminate) when these limits
are reached. Generally we prefer larger limits than the default `maxiter=50` of `nls()`,
but that may simply reflect our history of dealing with more difficult problems as we are
the tool-makers; users consult us when things go wrong.

@@

This is the infamous "singular gradient" termination message of `nls()`. A Google search of

@@

gets over 4000 hits that are spread over the years. 
In some cases this is due
to the failure of the simple finite difference approximation of the Jacobian in the
`numericDeriv()` function that is a part of `nls()`. `nlsr` can use analytic
derivatives, and we could import this functionality to the `nls()` code as an
improvement. See below in the section **Jacobian computation**.

A common source of the issue is that the Jacobian is very close
to singular for some values of the model parameters. In such cases we need to find
an alternative algorithm to the Gauss-Newton iteration of `nls()`. The most common
work-around is the Levenberg-Marquardt stabilization (see @Marquardt1963, @Levenberg1944, 
@jn77ima). Versions of this have been implemented in packages `minpack.lm` and `nlsr`.
and we have prepared experimental versions of `nls` replacements that can incorporate
stabilizations. Integration of such ideas with other code structures so that all the
features of `nls()` work properly has unfortunately proved difficult.

Let us consider an example which presents the infamous 'singular gradient' error message.
This was posted on StackExchange. 
*https://stats.stackexchange.com/questions/13053/singular-gradient-error-in-nls-with-correct-starting-values*

```{r SESingGrad}
reala = -3
realb = 5
realc = 0.5
realr = 0.7
realm = 1
x = 1:11 #x values; 11 timepoint data
#linear+exponential function
y=reala + realb * realr^(x - realm) + realc * x
testdat = data.frame(x, y)
strt <- list(a = -3, b = 5, c = 0.5, r = 0.7, m = 1)
jform <- y ~ a + b * r^(x - m) + c * x # Formula
linexp=try(nls(jform, data = testdat, start = strt, trace = F))
library(nlsr)
# Note singular values of Jacobian in rightmost column
linexp2 <- try(nlxb(jform, data = testdat, start = strt, trace = F))
linexp2
```

In the above, we see that the Jacobian is essentially singular as shown by its singular
values. Note that we DISPLAY the singular values in a single column for convenience in
the output. The values do NOT correspond to individual parameters, but are a property
of the whole problem.

@@

`nls()`, with the `numericDeriv()` function, computes the Jacobian as the "gradient"
attribute of the residual vector. This is implemented as a mix of R and C code, but
we have created a rather more compact version entirely in R.
Code project. See the companion document **Jacobian Calculations for nls()**
(?? need citation -- currently **DerivsNLS.pdf**).

The Jacobian used by `nlsr::nlxb()` is computed from analytic expressions, while that
of `nls()` (and, by extension, `minpack.lm::nlsLM`) will use a numerical approximation. 
This can lead to different behaviour:

@@

Note that the selfStart models in the `./src/library/stats/R/zzModels.R` file
provide the Jacobian in the "gradient" attribute of the "one-sided" formula that defines each
model, and these Jacobians may be the analytic forms. The `nls()` function, after computing the
"right hand side" or `rhs` of the residual, then checks to see if the "gradient" attribute is
defined, and, if not, uses `numericDeriv` to compute a Jacobian into that attribute. This code
is within the `nlsModel()` or `nlsModel.pliner()` functions. The use of analytic Jacobians
almost certainly contributes to the performance of `nls()` on selfStart models.

@@

`nls()` accepts an argument `subset`. Unfortunately, this acts through the mediation of
`model.frame` and is not clearly obvious in the source code files `/src/library/stats/R/nls.R` and 
`/src/library/stats/src/nls.C`.

@@

The typical default action, which can be seen by using the command `getOption("na.action")`
is `na.omit`. This option essentially presents computations with data with all observations
containing any missing values (i.e. any row of a data frame with an NA) omitted. 
`na.exclude` does much the same for computations, but keeps the rows with NA elements so 
that predictions are in the correct row position. We recommend that workers actually test 
output to verify behaviour is as wanted.

@@

@@

`model` is an argument to the `nls()` function, which is documented

@@

Indeed, the argument only gets used when `nls()` is about to return its result object, and the
element `model` is NULL unless the calling argument `model` is TRUE. (Using the same name could
be confusing.) However, the model frame is used within the function code in the form of the object
`mf`.

@@

Nevertheless, to provide compatible behaviour with `nls()`, competing programs 
need to ensure equivalent behaviour, but users should test that the 
operation is correct.

@@

Nonlinear estimation algorithms are almost all iterative and need a set of starting
parameters. `nls()` offers a special class of modeling functions called **selfStart** 
models. There are a number of these in base R (see list below) and others in R packages
such as CRAN package `nlraa` (@MiguezNLRAA2021), as well as the now-archived package
`NRAIA`. Unfortunately, the structure of **selfStart** codes
is such that the methods by which initial parameters are computed is entangled
with the particulars of the `nls()` code. Though there is a `getInitial()` function,
this is not easy to use to simply compute the initial parameter estimates, in part 
because it may call `nls()`. Such circular references are, in our view, dangerous.

In the example below, we show how the `SSlogis` selfStart function can generate a set
of initial parameters for a 3-parameter logistic curve. The form used by `SSlogis`
is $$ y \sim Asym/(1+exp((xmid-tt)/scal)) $$, but we show how the starting parameters
for this model can be transformed to those of 
another form of the model, namely, $$ y \sim b1/(1 + b2*exp(-b3*t)) $$.

Let us look at the actual code for `SSlogis()` in `R-devel/src/library/stats/R/zzModels.R`:

```
SSlogis <- selfStart(~ Asym/(1 + exp((xmid - input)/scal)),
    selfStart(
        function(input, Asym, xmid, scal)
        {
              .expr1 <- xmid - input
              .expr3 <- exp(.e2 <- .expr1/scal)
              .expr4 <- 1 + .expr3
              .value <- Asym/.expr4
              .actualArgs <- as.list(match.call()[c("Asym", "xmid", "scal")])
              if(all(vapply(.actualArgs, is.name, NA)))
              {
            		  .expr10 <- .expr4^2
                  .grad <- array(0, c(length(.value), 3L), list(NULL, c("Asym", "xmid", "scal")))
                  .grad[, "Asym"] <- 1/.expr4
		              .grad[, "xmid"] <- - (xm <- Asym * .expr3/scal/.expr10)
		              .grad[, "scal"] <- xm * .e2
                  dimnames(.grad) <- list(NULL, .actualArgs)
                  attr(.value, "gradient") <- .grad
              }
              .value
        },
        initial = function(mCall, data, LHS, ...) {
              xy <- sortedXyData(mCall[["input"]], LHS, data)
              if(nrow(xy) < 4) {
                  stop("too few distinct input values to fit a logistic model")
              }
              z <- xy[["y"]]
              ## transform to proportion, i.e. in (0,1) :
              rng <- range(z); dz <- diff(rng)
              z <- (z - rng[1L] + 0.05 * dz)/(1.1 * dz)
              xy[["z"]] <- log(z/(1 - z))		# logit transformation
              aux <- coef(lm(x ~ z, xy))
              pars <- coef(nls(y ~ 1/(1 + exp((xmid - x)/scal)),
                               data = xy,
                               start = list(xmid = aux[[1L]], scal = aux[[2L]]),
                               algorithm = "plinear", ...))
              setNames(pars [c(".lin", "xmid", "scal")],
                       mCall[c("Asym", "xmid", "scal")])
        },
        parameters = c("Asym", "xmid", "scal"))
```

This R function includes analytic expressions for the Jacobian ("gradient").
These could be be useful to R users, especially if documented. Moreover, we
wonder why the programmers have chosen to save so many quantities in "hidden"
variables, i.e., with names preceded by ".". These are then not displayed by the `ls()`
command, making them difficult to access.

In the event that a selfStart model is not available, `nls()` sets all the starting parameters 
to 1. This is, in our view, tolerable, but could possibly be improved by using a set of values
that are slightly different e.g., in the case of a model $$  y \sim a*exp(-b*x) + c*exp(-d*x)$$
it would be useful to have $b$ and $d$ values different so the Jacobian is not singular. Thus,
some sort of sequence like  1.0, 1.1, 1.2, 1.3 for the four parameters might be better and it
can be provided quite simply instead of all 1's.

```{r hobbsSSlogis}
weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
          38.558, 50.156, 62.948, 75.995, 91.972)
tt <- 1:12

NLSformula0 <- y ~ b1/(1+b2*exp(-b3*tt))
NLSformula <- y ~ SSlogis(tt, Asym, xmid, scal)
NLSformulax <- y ~ Asym/(1+exp((xmid-tt)/scal))
NLStestdata <- data.frame(y=weed, tt=tt) 
s0 <- getInitial(NLSformula, NLStestdata)
print(s0)
# We transform the parameters for the NLSformula0 model of original specification.
s1<-list(b1=s0[1], b2=exp(s0[2]/s0[3]), b3=1/s0[3])
print(as.numeric(s1))
# No actual improvement because nls() has been already used to get the starting values,
# but we do get SEs
hobblog<-nls(NLSformula0, data=NLStestdata, start=s1)
summary(hobblog)
deviance(hobblog)
# nls fails without selfStart -- singular gradient, even on stabilized formula
try(hobblogx<-nls(NLSformulax, data=NLStestdata))
# But Marquardt is able to get a solution easily
library(nlsr)
hobblogxx<-nlxb(NLSformulax, data=NLStestdata, start=c(Asym=1, xmid=1, scal=1))
hobblogxx
```

### selfStart models in base R

The following models are provided (in file ./src/library/stats/R/zzModels.R)

```
SSasymp         - asymptotic regression model
SSasympOff      - alternate formulation of asymptotic regression model with offset
SSasympOrig     - exponential curve through the origin to an asymptote
SSbiexp         - y ~ ~ A1 * exp(-exp(lrc1)*input) + A2 * exp(-exp(lrc2) * input)
SSfol           - y ~ Dose * (exp(lKe + lKa - lCl) * (exp(-exp(lKe) * input) -
                          exp(-exp(lKa) * input))/(exp(lKa) - exp(lKe)))
SSfpl           - four parameter logistic model
SSlogis         - three parameter logistic model
SSmicmen        - Michaelis-Menten model for enzyme kinetics
SSgompertz2     - Gompertz model for growth curve data
SSweibull       - Weibull model for growth curve data
```

>>>>

Based on work for a Google Summer of Code project "Improvements to `nls()`",
we consider the features and limitations of the R function `nls()`
in the context of improving and rationalizing R tools for nonlinear regression.

An important consideration is the  maintainability of
the code base that provides the functionality `nls()` claims to offer.
Our work suggests that the
existing code makes maintenance and improvement very difficult,
with legacy applications and examples blocking some important updates.
Discussion of these matters is relevant to improving R generally, as
well as its nonlinear estimation tools.

@@

`nls()` is the tool in base R (the distributed software package from [CRAN](https://cran.r-project.org)
for estimating nonlinear statistical models. The function dates to the 1980s and the work related
to @bateswatts in S (see https://en.wikipedia.org/wiki/S_%28programming_language%29).

The `nls()` function has a remarkable and comprehensive set of capabilities for estimating
nonlinear models that are expressed as **formulas**. In particular, we note that it

- handles formulas that include R functions
- allows data to be subset
- permits parameters to be indexed over a set of related data
- produces measures of variability (i.e., standard error estimates) for the estimated parameters
- has related profiling capabilities for exploring the likelihood surface as parameters are changed
- links to a number of pre-coded ("selfStart") models

With such a range of features and long history, the code has become untidy 
and overly patched. It is, to
our mind, essentially unmaintainable. Moreover, its underlying methods can and should be improved, 
as we suggest below.

@@

This amounts to admitting R cannot solve well-posed problems unless data is polluted
with errors.

This issue can be easily resolved. The "termination test" for the **program** rather than for 
"convergence" of the underlying **algorithm** is the Relative Offset Convergence Criterion 
(see @BatesWatts81). This projects the proposed step in the parameter vector on the gradient 
and estimates how much the sum of squares loss function should decrease. This estimate is divided 
by the current size of the loss function to avoid scale issues.
When we have "converged", the estimated
decrease is very small. However, with small residuals,
the sum of squares loss function is (almost) zero and we get the possibility of a 
zero-divide failure.

@@

<!-- ### Example of a small-residual problem -->

<!-- ``` -->
<!-- t <- -10:10 -->
<!-- y <- 100/(1 + .1 * exp(-0.51 * t)) -->
<!-- lform <- y ~ a/(1 + b * exp(-c * t)) -->
<!-- ldata <- data.frame(t = t, y = y) -->
<!-- plot(t, y) -->
<!-- lstartbad <- c(a = 1, b = 1, c = 1) -->
<!-- lstart2 <- c(a = 100, b = 10, c = 1) -->
<!-- nlsr::nlxb(lform, data = ldata, start = lstart2) -->
<!-- nls(lform, data = ldata, start = lstart2, trace = TRUE) -->
<!-- # Fix with scaleOffset -->
<!-- nls(lform, data = ldata, start = lstart2, trace = TRUE, control = list(scaleOffset = 1.0)) -->
<!-- sessionInfo() -->
<!-- ``` -->

<!-- Edited output of as small example follows: -->

<!-- ``` -->
<!-- > t <- -10:10 -->
<!-- > y <- 100/(1 + .1 * exp(-0.51 * t)) -->
<!-- > lform <- y ~ a/(1 + b * exp(-c * t)) -->
<!-- > ldata <- data.frame(t=t, y=y) -->
<!-- > plot(t,y) -->
<!-- > lstartbad <- c(a = 1, b = 1, c = 1) -->
<!-- > lstart2 <- c(a = 100, b = 10, c = 1) -->
<!-- > nlsr::nlxb(lform, data = ldata, start = lstart2) -->
<!-- nlsr object: x  -->
<!-- residual sumsquares =  1.007e-19  on  21 observations -->
<!--     after  13    Jacobian and  19 function evaluations -->
<!--   name            coeff          SE       tstat      pval      gradient    JSingval    -->
<!-- a                    100     2.679e-11  3.732e+12  1.863e-216  -6.425e-11       626.6   -->
<!-- b                    0.1      3.78e-13  2.646e+11  9.125e-196  -3.393e-08       112.3   -->
<!-- c                   0.51       6.9e-13  7.391e+11  8.494e-204   1.503e-08       2.791   -->
<!-- > nls(lform, data = ldata, start = lstart2, trace = TRUE) -->
<!-- 40346.    (1.08e+00): par = (100 10 1) -->
<!-- 11622.    (2.93e+00): par = (101.47 0.49449 0.71685) -->
<!-- 5638.0    (1.08e+01): par = (102.23 0.38062 0.52792) -->
<!-- 642.08    (1.04e+01): par = (102.16 0.22422 0.41935) -->
<!-- 97.712    (1.79e+01): par = (100.7 0.14774 0.45239) -->
<!-- 22.250    (1.78e+02): par = (99.803 0.093868 0.50492) -->
<!-- 0.025789  (1.33e+03): par = (100.01 0.10017 0.50916) -->
<!-- 6.0571e-08 (7.96e+05): par = (100 0.1 0.51) -->
<!-- 4.7017e-19 (1.86e+04): par = (100 0.1 0.51) -->
<!-- 1.2440e-27 (5.71e-01): par = (100 0.1 0.51) -->
<!-- 1.2440e-27 (5.71e-01): par = (100 0.1 0.51) -->
<!-- ... (approx 40 lines omitted) -->
<!-- 1.2440e-27 (5.71e-01): par = (100 0.1 0.51) -->
<!-- Error in nls(lform, data = ldata, start = lstart2, trace = TRUE) :  -->
<!--   number of iterations exceeded maximum of 50 -->
<!-- > # Fix with scaleOffset -->
<!-- > nls(lform, data = ldata, start = lstart2, trace = TRUE, control = list(scaleOffset = 1.0)) -->
<!-- 40346.      (1.08e+00): par = (100 10 1) -->
<!-- 11622.      (2.91e+00): par = (101.47 0.49449 0.71685) -->
<!-- 5638.0      (9.23e+00): par = (102.23 0.38062 0.52792) -->
<!-- 642.08      (5.17e+00): par = (102.16 0.22422 0.41935) -->
<!-- 97.712      (2.31e+00): par = (100.7 0.14774 0.45239) -->
<!-- 22.250      (1.11e+00): par = (99.803 0.093868 0.50492) -->
<!-- 0.025789    (3.79e-02): par = (100.01 0.10017 0.50916) -->
<!-- 6.0571e-08  (5.80e-05): par = (100 0.1 0.51) -->
<!-- 4.7017e-19  (1.62e-10): par = (100 0.1 0.51) -->
<!-- Nonlinear regression model -->
<!--   model: y ~ a/(1 + b * exp(-c * t)) -->
<!--    data: ldata -->
<!--      a      b      c  -->
<!-- 100.00   0.10   0.51  -->
<!--  residual sum-of-squares: 4.7e-19 -->
<!-- Number of iterations to convergence: 8  -->
<!-- Achieved convergence tolerance: 1.62e-10 -->
<!-- ``` -->

@@

The single convergence criterion of `nls()` leaves out some possibilities that could
be useful for some problems. The package `nlsr` (@nlsr-manual) already offers both 
the safeguarded relative offset test (**roffset**) as well as a **small sum of 
squares** test (**smallsstest**) that compares the latest evaluated sum of squared
(weighted) residuals to `e4` times the initial sum of squares, where
`e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.

We note that `nls()` uses a termination test to stop after `maxiter` "iterations". 
Unfortunately, the meaning of "iteration" varies among programs and requires 
careful examination of the
code. We prefer to use the number of times the residuals or the jacobian have
been computed and put upper limits on these. Our codes exit (terminate) when these limits
are reached. Generally we prefer larger limits than the default `maxiter=50` of `nls()`,
but that may reflect the more difficult problems we have encountered because 
users consult us when standard tools have given unsatisfactory results.

@@

This refers to the infamous "singular gradient" termination message of `nls()`. 
A Google search of

@@

gets over 4000 hits that are spread over some years. This could be because the 
Jacobian is poorly approximated (see **Jacobian computation** below). However,
it is common in nonlinear least squares computations that the Jacobian is very close
to singular for some values of the model parameters. In such cases we need to find
an alternative algorithm to the Gauss-Newton iteration of `nls()`. The most common
work-around is the Levenberg-Marquardt stabilization (see @Marquardt1963, @Levenberg1944, 
@jn77ima), and versions of it have been implemented in packages `minpack.lm` and `nlsr`.
In our work we prepared experimental prototypes of `nls` that incorporate
stabilizations, but integration with all the features of `nls()` has proved difficult.

Let us consider an 'singular gradient' example 
*https://stats.stackexchange.com/questions/13053/singular-gradient-error-in-nls-with-correct-starting-values*.
`nlsr::nlxb()` displays the singular values of the Jacobian, so the `nls()` error message is correct,
but unhelpful in obtaining a solution.

```{r SESingGrad}
reala = -3; realb = 5; realc = 0.5; realr = 0.7; realm = 1 # underlying parameters
x = 1:11 # x values; 11 data points
y=reala + realb * realr^(x - realm) + realc * x # linear+exponential function
testdat = data.frame(x, y) # save in a data frame
strt <- list(a = -3, b = 5, c = 0.5, r = 0.7, m = 1) # give programs a good start
jform <- y ~ a + b * r^(x - m) + c * x # Formula
library(nlsr)
linexp2 <- try(nlxb(jform, data = testdat, start = strt, trace = F))
linexp2 # Note singular values of Jacobian in rightmost column
# We get an error with nls()
linexp=try(nls(jform, data = testdat, start = strt, trace = F))
```

The Jacobian is essentially singular as shown by its singular
values. Note that these are **displayed** in a single column in
the output to provide a compact layout, but the values do **NOT** correspond to 
the individual parameters in 
whose row they appear; they are a property of the whole problem.

@@

`nls()`, with the `numericDeriv()` function, computes the Jacobian as the "gradient"
attribute of the residual vector. This is implemented as a mix of R and C code, but
we have created a rather more compact version entirely in R.
<!-- Code project. See the companion document **Jacobian Calculations for nls()** -->
<!-- (?? need citation -- currently **DerivsNLS.pdf**). -->

The Jacobian used by `nlsr::nlxb()` is computed from analytic expressions, while that
of `nls()` (and, by extension, `minpack.lm::nlsLM`) mostly use a numerical approximation. 
This can lead to different behaviour:

@@

Note that the selfStart models in the base R `./src/library/stats/R/zzModels.R` file
provide the Jacobian in the "gradient" attribute of the "one-sided" formula that defines each
model, and these Jacobians may be the analytic forms. The `nls()` function, after computing the
"right hand side" or `rhs` of the residual, then checks to see if the "gradient" attribute is
defined, and, if not, uses `numericDeriv` to compute a Jacobian into that attribute. This code
is within the `nlsModel()` or `nlsModel.pliner()` functions. The use of analytic Jacobians
almost certainly contributes to the good performance of `nls()` on selfStart models.

@@

`nls()` accepts an argument `subset`. This acts through the mediation of
`model.frame` and is not clearly obvious in the source code files `/src/library/stats/R/nls.R` and 
`/src/library/stats/src/nls.C`.

@@

The typical default action, which can be seen by using the command `getOption("na.action")`
is `na.omit`. This option essentially omits from computations any observations
containing missing values (i.e. any row of a data frame containing an NA). 
`na.exclude` does much the same for computations, but keeps the rows with NA elements so 
that predictions are in the correct row position. We recommend that workers actually test 
output to verify behaviour is as wanted.

@@

Our concern with `na.action` is that users may be unaware of the effects of an option
setting they may not even be aware has been set. Is `na.fail` a safer default?

@@

`model` is an argument to the `nls()` function, which is documented as

@@

Indeed, the argument only gets used when `nls()` is about to return its result object, and the
element `model` is NULL unless the calling argument `model` is TRUE. (Using the same name could
be confusing.) However, the model frame is used within the function code in the form of the object
`mf`. We feel that users could benefit from more extensive documentation and examples of its use.

@@

Nevertheless, to provide compatible behaviour with `nls()`, competing programs 
need to ensure equivalent behaviour, but users should test that the 
operation is as they intend.

@@

Nonlinear estimation algorithms are almost all iterative and need a set of starting
parameters. `nls()` offers a special class of modeling functions called **selfStart** 
models. There are a number of these in base R (`./src/library/stats/R/zzModels.R`)
and others in R packages such as CRAN package `nlraa` (@MiguezNLRAA2021), as well as 
the now-archived package `NRAIA`. Unfortunately, some **selfStart** codes
entangle the calculation of starting values for parameters
with the particulars of the `nls()` code. Though there is a `getInitial()` function,
this is not easy to use to simply compute the initial parameter estimates, in part 
because it may call `nls()`. Such circular references are, in our view, dangerous.

Other concerns are revealed by the example below. Here the `SSlogis` selfStart function 
is used to generate a set
of initial parameters for a 3-parameter logistic curve. The form used by `SSlogis`
is $y \sim Asym/(1+exp((xmid-tt)/scal))$, but we show how the starting parameters
for this model can be transformed to those of 
another form of the model, namely, $y \sim b1/(1 + b2*exp(-b3*t))$.

The code for `SSlogis()` is in `./src/library/stats/R/zzModels.R`.
<!-- ``` -->
<!-- SSlogis <- selfStart(~ Asym/(1 + exp((xmid - input)/scal)), -->
<!--     selfStart( -->
<!--         function(input, Asym, xmid, scal) -->
<!--         { -->
<!--               .expr1 <- xmid - input -->
<!--               .expr3 <- exp(.e2 <- .expr1/scal) -->
<!--               .expr4 <- 1 + .expr3 -->
<!--               .value <- Asym/.expr4 -->
<!--               .actualArgs <- as.list(match.call()[c("Asym", "xmid", "scal")]) -->
<!--               if(all(vapply(.actualArgs, is.name, NA))) -->
<!--               { -->
<!--             		  .expr10 <- .expr4^2 -->
<!--                   .grad <- array(0, c(length(.value), 3L), list(NULL, c("Asym", "xmid", "scal"))) -->
<!--                   .grad[, "Asym"] <- 1/.expr4 -->
<!-- 		              .grad[, "xmid"] <- - (xm <- Asym * .expr3/scal/.expr10) -->
<!-- 		              .grad[, "scal"] <- xm * .e2 -->
<!--                   dimnames(.grad) <- list(NULL, .actualArgs) -->
<!--                   attr(.value, "gradient") <- .grad -->
<!--               } -->
<!--               .value -->
<!--         }, -->
<!--         initial = function(mCall, data, LHS, ...) { -->
<!--               xy <- sortedXyData(mCall[["input"]], LHS, data) -->
<!--               if(nrow(xy) < 4) { -->
<!--                   stop("too few distinct input values to fit a logistic model") -->
<!--               } -->
<!--               z <- xy[["y"]] -->
<!--               ## transform to proportion, i.e. in (0,1) : -->
<!--               rng <- range(z); dz <- diff(rng) -->
<!--               z <- (z - rng[1L] + 0.05 * dz)/(1.1 * dz) -->
<!--               xy[["z"]] <- log(z/(1 - z))		# logit transformation -->
<!--               aux <- coef(lm(x ~ z, xy)) -->
<!--               pars <- coef(nls(y ~ 1/(1 + exp((xmid - x)/scal)), -->
<!--                                data = xy, -->
<!--                                start = list(xmid = aux[[1L]], scal = aux[[2L]]), -->
<!--                                algorithm = "plinear", ...)) -->
<!--               setNames(pars [c(".lin", "xmid", "scal")], -->
<!--                        mCall[c("Asym", "xmid", "scal")]) -->
<!--         }, -->
<!--         parameters = c("Asym", "xmid", "scal")) -->
<!-- ``` -->
This R function includes analytic expressions for the Jacobian ("gradient").
These could be be useful to R users, especially if documented. Moreover, we
wonder why the programmers have chosen to save so many quantities in "hidden"
variables, i.e., with names preceded by ".". These are then not displayed by the `ls()`
command, making them difficult to access.

In the event that a selfStart model is not available, `nls()` sets all the starting parameters 
to 1. This is, in our view, tolerable, but could be improved by using a set of values
that are all slightly different, which, in the case of the example model $y \sim a*exp(-b*x) + c*exp(-d*x)$
would avoid a singular Jacobiam because $b$ and $d$ were equal in value. 
A sequence like  1.0, 1.1, 1.2, 1.3 for the four parameters 
could be provided quite simply instead of all 1's.

<!-- ```{r hobbsSSlogis} -->
<!-- weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, -->
<!--           38.558, 50.156, 62.948, 75.995, 91.972) -->
<!-- tt <- 1:12 -->

<!-- NLSformula0 <- y ~ b1/(1+b2*exp(-b3*tt)) -->
<!-- NLSformula <- y ~ SSlogis(tt, Asym, xmid, scal) -->
<!-- NLSformulax <- y ~ Asym/(1+exp((xmid-tt)/scal)) -->
<!-- NLStestdata <- data.frame(y=weed, tt=tt)  -->
<!-- s0 <- getInitial(NLSformula, NLStestdata) -->
<!-- print(s0) -->
<!-- # We transform the parameters for the NLSformula0 model of original specification. -->
<!-- s1<-list(b1=s0[1], b2=exp(s0[2]/s0[3]), b3=1/s0[3]) -->
<!-- print(as.numeric(s1)) -->
<!-- # No actual improvement because nls() has been already used to get the starting values, -->
<!-- # but we do get SEs -->
<!-- hobblog<-nls(NLSformula0, data=NLStestdata, start=s1) -->
<!-- summary(hobblog) -->
<!-- deviance(hobblog) -->
<!-- # nls fails without selfStart -- singular gradient, even on stabilized formula -->
<!-- try(hobblogx<-nls(NLSformulax, data=NLStestdata)) -->
<!-- # But Marquardt is able to get a solution easily -->
<!-- library(nlsr) -->
<!-- hobblogxx<-nlxb(NLSformulax, data=NLStestdata, start=c(Asym=1, xmid=1, scal=1)) -->
<!-- hobblogxx -->
<!-- ``` -->

<!-- ### selfStart models in base R -->

<!-- The following models are provided (in file ./src/library/stats/R/zzModels.R) -->

<!-- ``` -->
<!-- SSasymp         - asymptotic regression model -->
<!-- SSasympOff      - alternate formulation of asymptotic regression model with offset -->
<!-- SSasympOrig     - exponential curve through the origin to an asymptote -->
<!-- SSbiexp         - y ~ ~ A1 * exp(-exp(lrc1)*input) + A2 * exp(-exp(lrc2) * input) -->
<!-- SSfol           - y ~ Dose * (exp(lKe + lKa - lCl) * (exp(-exp(lKe) * input) - -->
<!--                           exp(-exp(lKa) * input))/(exp(lKa) - exp(lKe))) -->
<!-- SSfpl           - four parameter logistic model -->
<!-- SSlogis         - three parameter logistic model -->
<!-- SSmicmen        - Michaelis-Menten model for enzyme kinetics -->
<!-- SSgompertz2     - Gompertz model for growth curve data -->
<!-- SSweibull       - Weibull model for growth curve data -->
<!-- ``` -->
